Hardware Caches and Optimization

### 2017-06-25 paper.tex
Tweak
Expand content, revise, tweak
Add TikZ graphic representing an array of pointers
Tweak float placement
Add some org-reveal slides; expand, revise, tweak
http://orgmode.org/
https://github.com/yjwen/org-reveal
https://github.com/hakimel/reveal.js/
Minor tweaks and additions
Tweak stuff and fill in some blanks

@@

% Prevent floats from moving into another section (https://ctan.org/pkg/placeins).
\usepackage[above, below, section]{placeins}

@@

% Stack for generating tables from CSV files.  Loading is done by [pgfplotstable].  It can
% also round, format and post-process data.  See [1].
% [siunitx] adds the `S` column type for advanced aligning in tables like around the
% decimal marker.  Getting this to work is a bit tricky; see [2].  There is some support
% for aligning around the decimal marker in [pgfplotstable] (see [3]), but I couldn't get
% it to work.  One can also right-align numbers relatively to each other but still center
% the column as a whole with [siunitx] (see [4]).
% [pgfplotstable]: https://ctan.org/pkg/pgfplotstable
% [siunitx]: https://ctan.org/pkg/siunitx
% [1]: https://tex.stackexchange.com/q/146716
% [2]: https://tex.stackexchange.com/q/131081
% [3]: https://tex.stackexchange.com/q/276485
% [4]: https://tex.stackexchange.com/q/9203
\usepackage{pgfplotstable}
\usepackage[binary-units]{siunitx} % Load `\kibi` and `\mebi` etc. (`binary-units`).
% These packages are recommended by [pgfplotstable].  The booktabs package adds the
% `\toprule`, `\midrule`, and `\bottomrule` commands for better looking tables.  See [5].
\usepackage{booktabs, array, colortbl}
% [booktabs]: https://ctan.org/pkg/booktabs
% [5]: https://en.wikibooks.org/wiki/LaTeX/Tables#Professional_tables

% This gets loaded by the `pgfplotstable` sub-package as far as I know.
%
% \usepackage{pgfplots}
%
% It then suggests:
%
%    Package pgfplots Warning: running in backwards compatibility mode (unsuitable t
%    ick labels; missing features). Consider writing \pgfplotsset{compat=1.14} into
%    your preamble
%
% So I guess I'll go that.
\pgfplotsset{compat=1.14}

@@

\usepackage{tikz}
\input{tex/datavis.tex}

@@

% Top matter {{{1
\title{Basics of Hardware Cache Optimization}
% \title{Algorithms for Hardware Caches}
% \title{Basics of CPU Cache Optimization}
% \title{Basics of Optimizing for Hardware Caches}
% \title{Fundamentals of Optimizing for Hardware Caches}
% \title{Introduction to Optimizations for Hardware Caches}
% \title{Optimizing for Hardware Caches}
% \subtitle{The Basics}
\author{Lukas Waymann}

@@

   The abstract \gls{emm} for memory hierarchies is explained and a small selection of
   algorithms \alts{developed for it, analyzed under it} are explored mathematically and
   empirically.
\end{abstract}
\newpage

@@

\input{tex/intro}
\input{tex/motivation}
% \section{Hardware Caches} % Cache, Structure of Caches
\input{tex/operation}
\input{tex/types}
\input{tex/key-concepts}
\input{tex/list-vs-vector.tex}
\label{sec:lvv}
% \section{Cache Replacement Policies} % Eviction Policies/Strategies/Algorithms
% https://en.wikipedia.org/wiki/Cache_replacement_policies
\input{tex/emm.tex}
\section{Parallel Computing}
\subsection{Cache Coherence} % Coherence or coherency?
% https://en.wikipedia.org/wiki/Cache_coherence
% Dirty.  Invalid.
\subsection{False Sharing}

>>>>

% Required for `\mathbb{}`.  TODO: `amssymb` loads `amsfonts`; which one should I use?
\usepackage{amsfonts}

% Used to put TikZ graphics into their own dedicated files and compile them separately.
% From the manual: "Load the standalonepackage very early in the main document."
\usepackage{standalone}

@@

% XXX: I used to specify the `section` option here to prevent floats from moving into
% another section.  It redefines the `\section` command to add a `\FloatBarrier`.  Turns
% out this doesn't work the way I expected.  Apparently, `\FloatBarrier` just issues
% `\clearpage` if there are floats queued [1][2].  The result can be mostly empty pages.
%
% [1]: https://tex.stackexchange.com/q/223917#comment525956_223917
% [2]: https://tex.stackexchange.com/q/272858#comment655658_272858
% [3]: https://tex.stackexchange.com/q/88657
% [4]: https://ctan.org/pkg/placeins)
\usepackage[above, below]{placeins}

% TODO: `\usepackage{afterpage}`?
%
% [1]: https://tex.stackexchange.com/q/88657
% [2]: https://www.ctan.org/pkg/afterpage

@@

\input{tex/pgfplotstablesetup}

@@

\usepackage{tikz}
\usetikzlibrary{math}
\input{tex/datavis}

@@

% Macro for typesetting "C++".  Just writing it verbatim looks very ugly.  See
% <https://stackoverflow.com/q/2724760>.
\usepackage{relsize} % For `\smaller`.
\def\ifmonospace{\ifdim\fontdimen3\font=0pt}
\def\cpp{%
\ifmonospace%
    C++%
\else%
    C\kern-.0667em\raise.30ex\hbox{\smaller{++}}%
\fi%
\spacefactor1000}
% Alternative from <https://tex.stackexchange.com/q/4302>:
% \def\cpp{{C\nolinebreak[4]\hspace{-.05em}\raisebox{.4ex}{\tiny\bfseries++}}}

% Top matter {{{1
\title{Hardware Caches and Optimization}
% \title{Optimizing for Hardware Caches}
% \title{Basics of Hardware Cache Optimization}
% \title{Algorithms for Hardware Caches}
% \title{Basics of CPU Cache Optimization}
% \title{Basics of Optimizing for Hardware Caches}
% \title{Fundamentals of Optimizing for Hardware Caches}
% \title{Introduction to Optimizations for Hardware Caches}
% \subtitle{The Basics}
\author{Lukas Waymann}

@@

   The abstract \gls{emm} for memory hierarchies and the \gls{com} derived from it are
   presented briefly.
\end{abstract}
\newpage

@@

\input{tex/intro}
\input{tex/motivation}
% TODO: "Hardware Caches", "Cache", or "Structure of Caches" section?
\input{tex/operation}
\input{tex/types}
\input{tex/key-concepts}
\input{tex/list-vs-vector}
% TODO: "Cache Replacement Policies", "Eviction (Policies|Strategies|Algorithms)" section?
% See <https://en.wikipedia.org/wiki/Cache_replacement_policies>.
\input{tex/abstract-}
% \input{tex/parallel-computing}
% \input{tex/false-sharing}
% TODO: "Roofline" section?  See <https://en.wikipedia.org/wiki/Roofline_model>.

### 2017-06-25 tex/types.tex
Add some org-reveal slides; expand, revise, tweak
http://orgmode.org/
https://github.com/yjwen/org-reveal
https://github.com/hakimel/reveal.js/
Add more slides, tweak stuff
Improve spacing between listings and captions
Tweak stuff and fill in some blanks

% Terminology / Nomenclature.
In practice, a \alts{currently, presently} representative%
\footnote{%
   % https://en.wikipedia.org/wiki/Bobcat_(microarchitecture)
   E.g. for AMD Family 14h processors~\cite[30--32]{14h},
   % https://en.wikipedia.org/wiki/List_of_AMD_CPU_microarchitectures
   % https://en.wikipedia.org/wiki/Zen_(microarchitecture)
   % 32 KiB L1d, 64 KiB L1i, 512 KiB L2, 8 to 16 MiB L3
   AMD Zen (17h)~\cite{zen}, and
   % https://en.wikipedia.org/wiki/Kaby_Lake
   % https://en.wikipedia.org/wiki/Skylake_(microarchitecture)
   % 32 KiB L1d, 32 KiB L1i, 256 KiB L2, 2 to 8 MiB L3
   Intel Skylake desktop processors%
   ~\cite[figure 2-1, table 2-4]{skylake}
   % ~\cite[figure 2-1, \pno~2-2, table 2-4, \pno~2-6]{skylake}.
   % ~\cite[{2-2}, {2-6}]{skylake}.
   % <https://en.wikipedia.org/wiki/Bulldozer_(microarchitecture)> is too weird.
}
x86 cache hierarchy consists of:
\begin{itemize}
   % https://en.wikipedia.org/wiki/Cache_hierarchy#Shared_versus_private
   \item Separate level 1 data and instruction caches of 32 to 64 KiB for each core
      (denoted \gls{l1d} and \gls{l1i} by  \textcite[14--15]{drepper2007}).
      % TODO?  Why have a separate instruction cache?
      Machine instructions in \gls{l1i} are already decoded%
      ~\cite[31, 56]{drepper2007}.
      % ~\cite[14, 31, 56]{drepper2007}.
   % \item A level 2 cache for \say{both code and data}~\cite[31]{drepper2007}.
   \item A unified \gls{l2} cache of 256 to 512 KiB for each core.
   \item Often a unified \gls{l3} cache of 2 to 16 MiB shared between all cores.
   \item TODO: Some \glspl{tlb} I guess.
\end{itemize}

@@

% XXX: hacks!  Use the figure environment so that LaTeX won't display this listing after
% the plot showing the results of profiling it.  "LaTeX [only] keeps all floats of the
% same *type* in order" [1].  Use `\captionof` to label the listing correctly as a
% listing.
%
% [1]: https://tex.stackexchange.com/q/127742/#comment290982_127744
\begin{comment}
   \begin{figure}
      \inputminted[firstline=9]{c}{access-times/access-times.c}
      \captionof{listing}{TODO}
      \label{lst:access-times}
   \end{figure}
\end{comment}

% \newenvironment{code}{\captionsetup{type=listing}}{}
% \begin{code}
%    \inputminted[firstline=9]{c}{access-times/access-times.c}
%    \caption{TODO}
%    \label{lst:access-times}
% \end{code}

@@

% Actually, don't use a float.  The listing should be allowed to span multiple pages which
% floats aren't.
%
% [1]: https://tex.stackexchange.com/q/14522/#comment484569_75880
% [2]: https://tex.stackexchange.com/q/175650
%      "How to allow page break inside a float environment?"
% [3]: https://tex.stackexchange.com/q/12428
\begin{center} % XXX: hack to get normal spacing after the caption and before the listing.
   \inputminted[firstline=12]{c}{access-times/access-times.c}
   % \captionof{listing}{}
   % \captionof{listing}{Randomly Access Elements of \mintinline{text}{array}}
   \captionof{listing}{Randomly Read Array Elements}
   % \captionof{listing}{Random Reads\label{lst:access-times}}
   \label{lst:access-times}
\end{center}

@@

\begin{figure}
   \centering
   \tikz \datavisualization[array size vs cycles plot]
      data [read from file=access-times/access-times.csv, separator=\space];
   % \caption{Access Times for Random Reads (\Cref{lst:access-times})}
   \caption{Access Times for Random Reads}
   \label{fig:access-times}
\end{figure}

% vim: tw=90 sts=-1 sw=3 et fdm=marker

>>>>

% Terminology / Nomenclature.
In practice, a \alts{currently, presently} representative%
\footnote{%
   % https://en.wikipedia.org/wiki/Bobcat_(microarchitecture)
   E.g. for AMD Family 14h processors~\cite[30--32]{14h},
   % https://en.wikipedia.org/wiki/List_of_AMD_CPU_microarchitectures
   % https://en.wikipedia.org/wiki/Zen_(microarchitecture)
   % 32 KiB L1d, 64 KiB L1i, 512 KiB L2, 8 to 16 MiB L3
   AMD Zen (17h)~\cite{zen}, and
   % https://en.wikipedia.org/wiki/Kaby_Lake
   % https://en.wikipedia.org/wiki/Skylake_(microarchitecture)
   % 32 KiB L1d, 32 KiB L1i, 256 KiB L2, 2 to 8 MiB L3
   Intel Skylake desktop processors%
   ~\cite[figure 2-1, table 2-4]{skylake}
   % ~\cite[figure 2-1, \pno~2-2, table 2-4, \pno~2-6]{skylake}.
   % ~\cite[{2-2}, {2-6}]{skylake}.
   % <https://en.wikipedia.org/wiki/Bulldozer_(microarchitecture)> is too weird.
}
x86 cache hierarchy consists of:
\begin{itemize}
   % https://en.wikipedia.org/wiki/Cache_hierarchy#Shared_versus_private
   \item Separate level 1 data and instruction caches of 32 to 64 KiB for each core
      (denoted \gls{l1d} and \gls{l1i} by  \textcite[14--15]{drepper2007}).
      % TODO?  Why have a separate instruction cache?
      Machine instructions in \gls{l1i} are already decoded%
      ~\cite[31, 56]{drepper2007}.
      % ~\cite[14, 31, 56]{drepper2007}.
   % \item A level 2 cache for \say{both code and data}~\cite[31]{drepper2007}.
   \item A unified \gls{l2} cache of 256 to 512 KiB for each core.
   \item Often a unified \gls{l3} cache of 2 to 16 MiB shared between all cores.
   \item One or more \glspl{tlb} per core.  \alts{They cache virtual-to-physical address
      associations of memory pages, They cache associations of virtual to physical
      addresses belonging to memory pages}.
\end{itemize}

@@

% XXX: hacks!  Use the figure environment so that LaTeX won't display this listing after
% the plot showing the results of profiling it.  "LaTeX [only] keeps all floats of the
% same *type* in order" [1].  Use `\captionof` to label the listing correctly as a
% listing.
%
% [1]: https://tex.stackexchange.com/q/127742/#comment290982_127744
\begin{comment}
   \begin{figure}
      \inputminted[firstline=9]{c}{access-times/access-times.c}%
      \captionof{listing}{TODO}
      \label{lst:access-times}
   \end{figure}
\end{comment}

% \newenvironment{code}{\captionsetup{type=listing}}{}
% \begin{code}
%    \inputminted[firstline=9]{c}{access-times/access-times.c}%
%    \caption{TODO}
%    \label{lst:access-times}
% \end{code}

@@

% Actually, don't use a float.  The listing should be allowed to span multiple pages which
% floats aren't.
%
% [1]: https://tex.stackexchange.com/q/14522/#comment484569_75880
% [2]: https://tex.stackexchange.com/q/175650
%      "How to allow page break inside a float environment?"
% [3]: https://tex.stackexchange.com/q/12428
\begin{center} % XXX: hack to get normal spacing after the caption and before the listing.
   \inputminted[firstline=12]{c}{access-times/access-times.c}%
   % \captionof{listing}{}
   % \captionof{listing}{Randomly Access Elements of \mintinline{text}{array}}
   \captionof{listing}{Randomly Read Array Elements}
   % \captionof{listing}{Random Reads\label{lst:access-times}}
   \label{lst:access-times}
\end{center}

@@

\begin{figure}
   \centering
   \input{tex/graphics/access-time-plot}
   \caption{Access Times for Random Reads}
   \label{fig:access-times}
\end{figure}

% vim: ft=tex tw=90 sts=-1 sw=3 et fdm=marker
